---
title: "Storm Data Analysis of National Weather Service Data"
output: 
  html_document:
    keep_md: true
---

######{r setup, include=FALSE}
######knitr::opts_chunk$set(echo = TRUE)


#                       Synopsis
#####    Storms and other severe weather events can cause both public health and economic problems for communities and municipalities.  
#####This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States.


#####    This project attemps to answer two questions:  
#####1) Across the United States, which types of events (as indicated in the "EVTYPE" variable) are most harmful with respect to population health?
#####2) Across the United States, which types of events have the greatest economic consequences?


#                       Data Processing
##### The data ("repdata_data_StormData.csv.bz2") was provided as part of the assignment documents and is included in the Github repo (https://github.com/GaryFH/RepDataStormStudy).  The data is "read" in, converted to a tbl_df and stored as "d1" with the following code:

```{r ReadData}
library(dplyr)
library(ggplot2)
c1<-read.csv("repdata_data_StormData.csv.bz2",stringsAsFactors = FALSE)
d1<-tbl_df(c1)
```
##### The data is initially examined with the following code:

```{r InitialLook}
d1
```
##### 985 different types of weather events - future investigations may want to look and combining some of these events (i.e all heat related events) The following code produces a plot that shows the top eight weather events based on fatalities.

```{r fatality}
d2<-group_by(d1,EVTYPE)
d3<-select(d2,EVTYPE,FATALITIES,INJURIES)
d4<-filter(d3,FATALITIES,!is.na(FATALITIES))
d5<-summarise(d4,FATALITIES=sum(FATALITIES))
d6<-arrange(d5,desc(FATALITIES))
d7<-head(d6,8)
g<-ggplot(d7,aes(EVTYPE,FATALITIES))
plotfatal<-g+geom_col(fill="yellow")+ geom_text(aes(label=FATALITIES), vjust=1.5, color="black")+labs(x="Weather event type" ,y="Total Fatalities",title="Top eight most dangerous weather events")
plotfatal
```
#### Notes
```{r property damage}
d2<-group_by(d1,EVTYPE)
dd3<-select(d2,EVTYPE,PROPDMG:CROPDMGEXP)
dd4<-filter(dd3,PROPDMG,!is.na(PROPDMG))
dd5<-summarise(dd4,PROPDMG=sum(PROPDMG)/100000)
dd6<-arrange(dd5,desc(PROPDMG))
dd7<-head(dd6,8)
dd7
g1<-ggplot(dd7,aes(EVTYPE,PROPDMG))
plotdamage<-g1+geom_col(fill="orange")+ geom_text(aes(label=PROPDMG), vjust=1.5)+labs(x="Weather event type",y="Total Property damage in billions",title="Top eight most expensive weather events")
plotdamage
print(plotdamage,fig.height=4)
```




