---
title: "Storm Data Analysis of National Weather Service Data"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#                       Synopsis
#####    Storms and other severe weather events can cause both public health and economic problems for communities and municipalities.  
#####This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States.


#####    This project attemps to answer two questions:  
#####1) Across the United States, which types of events (as indicated in the "EVTYPE" variable) are most harmful with respect to population health?
#####2) Across the United States, which types of events have the greatest economic consequences?


#                       Data Processing
##### The data ("repdata_data_StormData.csv.bz2") was provided as part of the assignment documents and is included in the Github repo (https://github.com/GaryFH/RepDataStormStudy).  The data is "read" in, converted to a tbl_df and stored as "d1" with the following code:

```{r ReadData}
library(dplyr)
library(ggplot2)
c1<-read.csv("repdata_data_StormData.csv.bz2",stringsAsFactors = FALSE)
d1<-tbl_df(c1)
```
##### The data is initially examined with the following code:

```{r InitialLook}
d1
```
##### 985 different types of weather events - future investigations may want to look and combining some of these events (i.e all heat related events) The following code produces a plot that shows the top eight weather events based on fatalities.

```{r fatality}
d2<-group_by(d1,EVTYPE)
d3<-select(d2,EVTYPE,FATALITIES,INJURIES)
d4<-filter(d3,FATALITIES,!is.na(FATALITIES))
d5<-summarise(d4,FATALITIES=sum(FATALITIES))
d6<-arrange(d5,desc(FATALITIES))
d7<-head(d6,8)
g<-ggplot(d7,aes(EVTYPE,FATALITIES))
plotfatal<-g+geom_col(fill="yellow")+ geom_text(aes(label=FATALITIES), vjust=1.5, color="black")

print(plotfatal,fig.width=8,fig.height=4)
```



