---
title: "Storm Data Analysis of National Weather Service Data"
output: 
  html_document:
    keep_md: true
---

######{r setup, include=FALSE}
######knitr::opts_chunk$set(echo = TRUE)


#                       Synopsis
#####    Storms and other severe weather events can cause both public health and economic problems for communities and municipalities.  
#####This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States.


#####    This project attemps to answer two questions:  
#####1) Across the United States, which types of events (as indicated in the "EVTYPE" variable) are most harmful with respect to population health?
#####2) Across the United States, which types of events have the greatest economic consequences?


#                       Data Processing
##### The data ("repdata_data_StormData.csv.bz2") was provided as part of the assignment documents and is included in the Github repo (https://github.com/GaryFH/RepDataStormStudy).  The data is "read" in, converted to a tbl_df and stored as "d1" with the following code:

```{r ReadData}
library(dplyr)
library(ggplot2)
c1<-read.csv("repdata_data_StormData.csv.bz2",stringsAsFactors = FALSE)
d1<-tbl_df(c1)
```
#### The data is initially examined with the following code:

```{r InitialLook}
d1
```


#### There are 985 different types of weather events in the data - future investigations may want to look and combining some of these events (i.e all heat related events, Hurricane&Hurricane/Typhoon and/or flood&storm-surge, related events) to see additional patterns. The following code produces a plot that shows the top eight weather events based on fatalities.

```{r fatality}
d2<-group_by(d1,EVTYPE)
d3<-select(d2,EVTYPE,FATALITIES,INJURIES)
d4<-d3##d4<-filter(d3,FATALITIES,!is.na(FATALITIES))
d5<-summarise(d4,FATALITIES=sum(FATALITIES),INJURIES=sum(INJURIES))
d6<-arrange(d5,desc(FATALITIES))
d7<-head(d6,8)
d7
g<-ggplot(d7,aes(x=reorder(EVTYPE,-FATALITIES),y=FATALITIES))
plotfatal<-g+geom_col(fill="yellow")+theme(axis.text.x = element_text(angle = 90, hjust = 1))+geom_text(aes(label=FATALITIES), vjust=1.5, color="black")+labs(x="Weather Event",title="Eight most Deadly Weather Events US 1950-2011")
plotfatal
```



#### The following code shows the top eight weather events based on injuries.

```{r injury}
g2<-ggplot(d7,aes(x=reorder(EVTYPE,-INJURIES),y=INJURIES))
plotinjury<-g2+geom_col(fill="magenta")+labs(x="Weather Event",title="Eight most Injuries due to Weather Events US 1950-2011")+theme(axis.text.x = element_text(angle = 90, hjust = 1))+ geom_text(aes(label=INJURIES), vjust=1.5, color="black")
plotinjury

```



#### The original dataset showing damage amounts in dollars are stored in two variables (PROPDMG & CROPDMG) the values in these columns need to be multiplied by factors that vary per observation - the code below adds two new columns(property damage multiplier - "pdmltplyr" & crop damage multiplier - "cdmltplyr") that make the character codes found in the dataset into numeric factors for each observation.

```{r clean damage data}
dd2<-group_by(d1,EVTYPE)
dd25<-mutate(dd2, pdmultplyr = 
        ifelse(PROPDMGEXP == "h" | PROPDMGEXP == "H", 100,
        ifelse(PROPDMGEXP == "k" | PROPDMGEXP == "K", 1000,
        ifelse(PROPDMGEXP == "m" | PROPDMGEXP == "M", 1000000,
        ifelse(PROPDMGEXP == "b" | PROPDMGEXP == "B", 1000000000, 0 )))))
dd26<-mutate(dd25, cdmultplyr=
        ifelse(CROPDMGEXP == "h" | CROPDMGEXP == "H", 100,
        ifelse(CROPDMGEXP == "k" | CROPDMGEXP == "K", 1000,
        ifelse(CROPDMGEXP == "m" | CROPDMGEXP == "M", 1000000,
        ifelse(CROPDMGEXP == "b" | CROPDMGEXP == "B", 1000000000, 0 )))))
dd3<-select(dd26,EVTYPE,PROPDMG:CROPDMGEXP,pdmultplyr,cdmultplyr)
```



####  The resulting "improved" dataframe can now be used to answer the question - which weather events caused the most damage?   The following code applys:

```{r, property damage}

dd35<-mutate(dd3,TotalDamage=(PROPDMG*pdmultplyr)+(CROPDMG*cdmultplyr))

dd4<-filter(dd35,TotalDamage,!is.na(TotalDamage))

dd5<-summarise(dd4,TotalDamage=sum(TotalDamage)/1000000000)
dd6<-arrange(dd5,desc(TotalDamage))
dd7<-head(dd6,8)
dd7
g3<-ggplot(dd7,aes(x=reorder(EVTYPE,-TotalDamage),y=TotalDamage))
plotdamage<-g3+geom_col(fill="orange")+labs(x="Weather Event",title="Eight highest damages due to Weather Events US 1950-2011")+ theme(axis.text.x = element_text(angle = 90, hjust = 1))+geom_text(aes(label=TotalDamage), vjust=1.5)+labs(x="Weather event type",y="Total Property damage in billions",title="Top eight most expensive weather events")
plotdamage
```


#                       SUMMARY

#### Clearly Tornados are the most dangerous weather events per this data set - Tornadoes have the most injuries and most fatalities by far.

#### Tornados also rank 3rd under most costly weather events behind both Hurricans/Typhoons(#2) and Floods(#1). There are 985 different types of weather events in the data - future investigations may want to look and combining some of these events (i.e all heat related events, Hurricane&Hurricane/Typhoon and/or flood&storm-surge, related events) to see additional patterns.

